---
title: "Functional Data Analysis of Spectrometric Data"
author: "Ketrin Hristova"
date: "2024-08-01"
output:
  pdf_document: default
  html_document: default
---

# Introduction

This project explores the application of Functional Data Analysis (FDA) on spectrometric data from meat samples. The goal is to analyze absorbance spectra and understand how these spectral features relate to the chemical composition of meat, specifically focusing on fat, water, and protein content.

The dataset used is the Tecator dataset, which includes 215 samples with spectrometric data and corresponding chemical compositions (fat, water, and protein content). The data consists of absorbance measurements across various wavelengths.

## Loading the Data and Checking the Structure

```{r}
library(fda.usc)
```

```{r}
data(tecator)
```

```{r}
str(tecator)
```

## Exploratory Data Analysis (EDA)

```{r}
library(ggplot2)

sample_indices <- sample(1:215, 10) 

matplot(tecator$absorp.fdata$argvals, t(tecator$absorp.fdata$data[sample_indices, ]), type = 'l',
xlab = "Wavelength (nm)",
ylab = "Absorbance",
main = "Random Spectrometric Curves",
col = 1:10,
lty = 1)
legend("topright", legend = paste("Sample", sample_indices), col = 1:10, lty
= 1, cex = 0.8)


```

# Plot histograms of the chemical components
```{r}
par(mfrow = c(1, 3)) 
# Fat content distribution
hist(tecator$y$Fat,
main = "Fat Content Distribution", xlab = "Fat (%)",
col = "lightblue",
border = "white")
# Water content distribution
hist(tecator$y$Water,
main = "Water Content Distribution", xlab = "Water (%)",
col = "lightgreen",
border = "white")
# Protein content distribution
hist(tecator$y$Protein,
main = "Protein Content Distribution", xlab = "Protein (%)",
col = "lightcoral",
border = "white")
```

## 1. Smoothing Functional Data


```{r}
library(fda)
library(fda.usc)
  dim(tecator$absorp.fdata$data)
```


```{r}
# Extract data for selected samples
original_data_samples <- tecator$absorp.fdata$data[sample_indices, ] 

# Original Curves for selected samples
matplot(tecator$absorp.fdata$argvals,
        t(original_data_samples),
        type = 'l',
        col = 1:length(sample_indices),
        lty = 1,
        lwd = 2,
        xlab = "Wavelength (nm)",
        ylab = "Absorbance",
        main = "Original Spectrometric Curves")

legend("topright", legend = paste("Sample", sample_indices), col =
1:length(sample_indices), lty = 1, lwd = 2)
  
```

# Apply B-spline Smoothing and Fourier Smoothing
```{r}
library(mgcv)
library(nlme)
set.seed(123) 
library(fda) 

tecator_transposed <- t(tecator$absorp.fdata$data)
# B-spline basis
n_basis <- 20
b_spline_basis <- create.bspline.basis(rangeval = tecator$absorp.fdata$rangeval, nbasis = n_basis)

# Smooth the data using B-splines
bspline_fd <- smooth.basis(tecator$absorp.fdata$argvals, tecator_transposed, b_spline_basis)$fd

# Fourier basis
fourier_basis <- create.fourier.basis(rangeval = tecator$absorp.fdata$rangeval, nbasis = 20)

# Smooth the data using Fourier series
fourier_fd <- smooth.basis(tecator$absorp.fdata$argvals, tecator_transposed, fourier_basis)$fd

par(mfrow = c(1, 1), mar = c(5, 5, 4, 2) + 0.1, cex.main = 1.5, cex.lab = 1.4, cex.axis = 1.2)

# B-spline smoothed curves for selected samples
matplot(tecator$absorp.fdata$argvals, 
        eval.fd(tecator$absorp.fdata$argvals, bspline_fd[sample_indices,]), 
        type = 'l',
        col = 1:length(sample_indices),
        lty = 1,
        lwd = 2,
        xlab = "Wavelength (nm)",
        ylab = "Absorbance",
        main = "B-spline Smoothed Curves")
legend("topright", legend = paste("Sample", sample_indices), col =
1:length(sample_indices), lty = 1, lwd = 2)
```

```{r}
# Fourier smoothed curves for selected samples
matplot(tecator$absorp.fdata$argvals, 
        eval.fd(tecator$absorp.fdata$argvals, fourier_fd[sample_indices,]), 
        type = 'l',
        col = 1:length(sample_indices),
        lty = 1,
        lwd = 2,
        xlab = "Wavelength (nm)",
        ylab = "Absorbance",
        main = "Fourier Smoothed Curves")
legend("topright", legend = paste("Sample", sample_indices), col =
1:length(sample_indices), lty = 1, lwd = 2)
```


```{r}
nbasis <- 30  

# Recreate the B-spline basis
ybasis <- create.bspline.basis(rangeval = tecator$absorp.fdata$rangeval, nbasis = nbasis)

# Optimize lambda using GCV with a lower-order derivative (2nd order)
loglam = seq(-6, 0, by = 0.25)
gcvsave = numeric(length(loglam))

for (i in 1:length(loglam)) {
  lambdai = 10^loglam[i]
  yfdPari = fdPar(ybasis, Lfdobj = int2Lfd(2), lambda = lambdai)  
  s = smooth.basis(tecator$absorp.fdata$argvals, tecator_transposed, yfdPari)
  
  gcvsave[i] = mean(s$gcv)
}

plot(loglam, gcvsave, type = 'l', xlab = 'log(lambda)', ylab = 'GCV', main = "GCV for B-spline Smoothing")

# Identify the optimal lambda
optimal_lambda_bspline = 10^loglam[which.min(gcvsave)]
cat("Optimal lambda for B-spline smoothing:", optimal_lambda_bspline, "\n")

# Refit with optimal lambda for B-spline
yfdPar_optimal_bspline = fdPar(ybasis, Lfdobj = int2Lfd(2), lambda = optimal_lambda_bspline)
smoothed_curves_optimal_bspline = smooth.basis(tecator$absorp.fdata$argvals, tecator_transposed, yfdPar_optimal_bspline)
smoothed_fd_optimal_bspline = smoothed_curves_optimal_bspline$fd

# Plot the optimally smoothed B-spline curves for selected samples
sample_indices <- c(1, 10, 50, 100, 150)  
smoothed_data_samples_optimal_bspline = eval.fd(tecator$absorp.fdata$argvals, smoothed_fd_optimal_bspline)[, sample_indices]

matplot(tecator$absorp.fdata$argvals, smoothed_data_samples_optimal_bspline, type = 'l', col = 1:length(sample_indices), lty = 1, lwd = 2, 
        xlab = "Wavelength (nm)", ylab = "Absorbance", main = "Optimally Smoothed B-spline Curves")
legend("topright", legend = paste("Sample", sample_indices), col = 1:length(sample_indices), lty = 1, lwd = 2)

```

```{r}

nbasis <- 30

# Recreate the B-spline basis
ybasis <- create.bspline.basis(rangeval = range(tecator$absorp.fdata$argvals), nbasis = nbasis)

# Use the optimal lambda value
optimal_lambda_bspline <- 0.3162278

# Create the functional parameter object using a lower order derivative
Lfdobj <- int2Lfd(2)  # Using the 2nd derivative instead of 4th
yfdPar_optimal_bspline <- fdPar(ybasis, Lfdobj = Lfdobj, lambda = optimal_lambda_bspline)

# Refit the data with the optimal lambda
smoothed_curves_optimal_bspline <- smooth.basis(tecator$absorp.fdata$argvals, tecator_transposed, yfdPar_optimal_bspline)
bspline_fd <- smoothed_curves_optimal_bspline$fd
```


```{r}
# Reset plot area
par(mfrow = c(1, 1))
```

## 3.Functional PCA

# Functional PCA on B-spline Smoothed Data
```{r}
library(fda)

pca_bspline <- pca.fd(bspline_fd, nharm = 5) 

# principal component scores and basis functions
pc_scores_bspline <- pca_bspline$scores
pc_basis_bspline <- pca_bspline$harmonics
```


# Principal Components
```{r}
par(mfrow = c(1,1), mar = c(5, 5, 4, 2) + 0.1, cex.main = 1.5, cex.lab = 1.4, cex.axis = 1.2)

# Principal Components for B-spline Smoothed Data
plot(pca_bspline$varprop[1:5], type = 'b', xlab = "Component", ylab = "Variance Explained",
     main = "Variance Explained by PCA (B-spline)", pch = 19)
```

```{r}
# Principal Component Functions for B-spline Smoothed Data
matplot(tecator$absorp.fdata$argvals, eval.fd(tecator$absorp.fdata$argvals, pc_basis_bspline),
        type = 'l', col = 1:5, lty = 1, lwd = 2,
        xlab = "Wavelength (nm)", ylab = "Principal Components", 
        main = "Principal Components (B-spline)")
legend("topright", legend = paste("PC", 1:5), col = 1:5, lty = 1, lwd = 2)
```



```{r}
# Reset plot area
par(mfrow = c(1, 1))
```

# Principal Component Scores

```{r}
# Principal Component Scores for B-spline Smoothed Data
print("Principal Component Scores for B-spline Smoothed Data:") 

print(head(pca_bspline$scores))
```

```{r}
fat_content <- tecator$y$Fat

# Compute the correlations between Fat content and the principal components
correlations <- cor(fat_content, pc_scores_bspline)

# Print the correlations
print("Correlations between Fat content and Principal Components:")
print(correlations)
```

```{r}
barplot(correlations, 
        names.arg = paste("PC", 1:length(correlations), sep=""), 
        col = "skyblue", 
        border = "black", 
        ylim = c(min(correlations) - 0.1, max(correlations) + 0.1),
        main = "Correlations between Fat Content and Principal Components",
        xlab = "Principal Components", 
        ylab = "Correlation with Fat Content")

# Add a horizontal line at y = 0 for reference
abline(h = 0, col = "black", lwd = 1)
```



# Variance Explained
```{r}
# Variance Explained for B-spline Smoothed Data
print("Variance Explained by Principal Components (B-spline):")

print(pca_bspline$varprop)
```




## 5. Depth Measures
```{r}
plot(bspline_fd)
```

```{r}
# boxplot using the Modified Band Depth (MBD) method
b1 <- boxplot(bspline_fd, method = "MBD")
```

```{r}
# boxplot using the BD2 method
b2 <- boxplot(bspline_fd, method = "BD2")
```

```{r}
# boxplot using the Both method (BD2 first, then MBD)
b3 <- boxplot(bspline_fd, method = "Both")
```


```{r}
print(b1$medcurve) 

print(b2$medcurve)

print(b3$medcurve) 

# output of the MBD method to extract the band depths
DM <- b1$depth 
summary(DM)
```

```{r}
# Histogram of depth measures
hist(DM, main = "Histogram of Functional Depth Measures", xlab = "Depth",
breaks = 30)
```

```{r}
# Boxplot of depth measures
boxplot(DM, main = "Boxplot of Functional Depth Measures", ylab = "Depth")
```

```{r}
# Calculate the quartiles and IQR
Q1 <- quantile(DM, 0.25) 
Q3 <- quantile(DM, 0.75) 
IQR <- Q3 - Q1

outlier_threshold <- Q1 - 1.5 * IQR 
outliers <- which(DM < outlier_threshold)

print("Outliers:") 

median_depth <- median(DM)
central_samples <- which(DM >= median_depth)

print("Central Samples:") 
print(central_samples)
```

```{r}
# Extract the depth measures (from MBD)
DM <- b1$depth

# Group the data into high-fat and low-fat groups
threshold <- 20  # Fat content threshold
index_high_fat <- which(tecator$y$Fat > threshold)
index_low_fat <- which(tecator$y$Fat <= threshold)

# Depth measures for high-fat and low-fat samples
D_high_fat <- DM[index_high_fat]
D_low_fat <- DM[index_low_fat]

# Perform Wilcoxon rank-sum test
wilcox_test_result <- wilcox.test(D_high_fat, D_low_fat)

# Print the test results
cat("Wilcoxon Rank-Sum Test Result:\n")
print(wilcox_test_result)
```


## Functional linear model with basis representation


```{r}
library(fda.usc)
data(tecator)

absorp <- tecator$absorp.fdata
fat20 <- ifelse(tecator$y$Fat < 20, 0, 1) * 2 + 2

par(mfrow = c(1, 3))
fat15 <- ifelse((y <- tecator$y$Fat) < 20, 2, 4)
boxplot(y, main = "Fat")
plot((X <- tecator$absorp), col = fat15, main = "Spectrometric: X")
plot((X.d1 <- fdata.deriv(tecator$absorp, 1)), col = fat15, main = "Derviative: X.d1")
```


```{r}
# first derivative of the absorbance spectra
absorp.d1 <- fdata.deriv(absorp, nderiv = 1)

# subset of the data
ind <- 1:165
tt <- absorp[["argvals"]]
y <- tecator$y$Fat[ind]
X <- absorp[ind, ]
X.d1 <- fdata.deriv(X, nbasis = 30, nderiv = 1)
X.d2 <- fdata.deriv(X, nbasis = 30, nderiv = 2)

rangett <- absorp$rangeval
basis1 <- create.bspline.basis(rangeval = rangett, nbasis = 30)

# Fit functional regression model using the first derivative
res.basis1 <- fregre.basis(X.d1, y, basis.x = basis1)

# Summary of the fitted functional regression model
print("Summary of Functional Regression Model:")
summary_res <- summary(res.basis1)
print(summary_res)

print("Detailed Output of Functional Regression Model:")
print(res.basis1)
```


```{r}
plot(y, res.basis1$fitted.values, main = "Functional Regression: Fitted vs Actual", 
     xlab = "Actual Fat Content", ylab = "Fitted Fat Content")
abline(0, 1, col = "red", lwd = 2)  # Add a 45-degree line for reference

plot(y, res.basis1$fitted.values, main = "Functional Regression with Linear Fit", 
     xlab = "Actual Fat Content", ylab = "Fitted Fat Content")
abline(lm(res.basis1$fitted.values ~ y), col = "blue", lwd = 2)  # Add linear regression line
abline(0, 1, col = "red", lwd = 2)  # Add a 45-degree line for reference

```






